Climate variability indices for ecological and crop models in R: the `climatrends` package

### 2020-06-30 inst/paper/paper.Rmd
paper

---
title: 'Climate variability indices for ecological and crop models in R: the `climatrends` package'
tags:
- climate data
- climatology
- earth science
- evapotranspiration
- precipitation data
- R
- reproducibility
- weather data
authors:
  - name: Kauê de Sousa
    orcid: 0000-0002-7571-7845
    affiliation: "1, 2"
  - name: Jacob van Etten
    orcid: 0000-0001-7554-2558
    affiliation: 2
  - name: Svein Ø. Solberg
    orcid: 0000-0002-4491-4483
    affiliation: 1
affiliations:
  - name: Department of Agricultural Sciences, Inland Norway University of Applied Sciences, Hamar, Norway
    index: 1
  - name: Bioversity International, Rome, Italy
    index: 2
citation_author: de Sousa et. al.
date: "`r format(Sys.time(), '%d %B %Y')`"
year: 2020
bibliography: paper.bib
csl: citationstyle.csl
output:
  pdf_document:
    keep_tex: true
header-includes: \usepackage{caption} \captionsetup[figure]{labelformat=empty}
---

```{r get data, include=FALSE}
library("knitr")
library("kableExtra")
library("tidyverse")
library("climatrends")
library("PlackettLuce")
funtbl <- read.csv("available_functions.csv")
load("cbean.rda")
```

@@

Abiotic factors plays an important role in most ecological and crop systems that depends on certain levels of temperature, light and precipitation (and their interplay) to initiate important physiological events [@PlantEcology]. In the walk of climate change, understand how these factors drives the physiological processes is a key approach to provide recommendations for adaptation and biodiversity conservation.

@@

`climatrends` aims to provide the R [@RCoreTeam] toolkit to compute extreme precipitation and temperature indices that serves as input for climate and crop models [@vanEtten2019; @Kehel2016], trends in climate change [@Aguilar2005; @deSousa2018] and applied ecology [@Prentice1992; @YLiu2018].

@@

Six main functions are provided, `crop_sensitive()`, `ETo()`, `GDD()`, `late_frost()`, `rainfall()` and `temperature()` with a default method for numeric 'vector's and additional methods implemented via the package `methods` [@RCoreTeam] for classes 'matrix' (or array), 'data.frame', and 'sf' (of geometry POINT or POLYGON) [@sf]. The later two are designed to fetch data from cloud sources.

@@

Some explanation on the `temperature()` and `rainfall()` functions.

@@

Where $T_{max}$ is the maximum temperature in the given day, $T_{min}$ is the minimum temperature in the given day and $T_{0}$ is the minimum temperature for growth (as per the physiology of the focal organism or ecosystem averages).

Additionally, the function `GDD()` offers three modified equations designed for cold environments and for tropical environments. For cold environments, where $T_{min}$ may be lower than $T_{0}$, there are two modified equations that adjusts either $T_{mean}$ (variant a) or $T_{min}$ (variant b). The variant a changes $T_{mean}$ to $T_{0}$ if $T_{mean} < T_{0}$ and is expressed as follows.

@@

Where $T$ may refer to $T_{min}$ and/or $T_{max}$ when the condition of being below $T_{0}$ applies.

@@

Where $T_{0_{max}}$ is the maximum base temperature for growth, defined in `GDD()` using the argument `tbase_max`.

@@

@@

Explain how the `late_frost()` is implemented.

@@

Two functions in **climatrends** are mainly designed to capture the effects of climate on the development and stress of crop species, `crop_sensitive()` computes indices that aims to capture the changes in temperature extremes during key phenological stages (e.g. anthesis), and `ETo()` computes the reference evapotranspiration.

The reference evapotranspiration measures the influence of the climate on a given plant water needs [@Brouwer1986]. The function `ETo()` applies the Blaney-Criddle method, a general theoretical method used when only air-temperature is available locally. It should be noted that this method is not very accurate and aims to provide the order of magnitude of evapotranspitation. The reference evapotranspiration is calculated using the following equation.

@@

Use the data from seed germination or crop growth to compute GDD. Maybe some GDD analysis in Scandinavia with data from the reindeer project.

## Common beans

During five growing seasons (from 2015 to 2017 in Nicaragua), van Etten et al. [-@vanEtten2019] conducted a crowdsourcing citizen-science trial testing 11 common beans varieties (*Phaseolus vulgaris* L.) in 842 farmer-managed experimental plots. Sets of three varieties were allocated randomly to farms as incomplete blocks. A Plackett–Luce model was used to analyse the data, this model estimates for each variety the probability that it wins, beating all other varieties in the set [@Turner2020]. An earlier version of `climatrends` was used in this research to capture the seasonal climate variation, here we reproduce part of this analysis regarding calculation and application of the climate indices. The approach here is slightly different because if considers the growing-degree days from planting date to maturity (the earlier study used planting date to the end of reproductive stage) and add new indices to illustrate the package implementation.

The data used here is available as supplementary materials (**zenodo ref**) as `cbean`. This contains a data.frame with a Plackett-Luce grouped rankings, the geographical coordinates of each sampled plot and the planting dates from where each farmer decided to start the experiment. The dates can differ from each other in the same season. The temperature data used was the land surface temperature MODIS (MYD11A2) [@Wan2015] and is storaged as an array with two layers (1st for the day and 2nd for the night temperatures). Each column corresponds to the dates (from 2015-09-10 to 2017-06-09) and the rows corresponds to the rows in the `cbean` data.frame.

@@

```{r table1, echo=FALSE, message=TRUE}
kable(funtbl, 
      row.names = FALSE,
      align = "l", 
      caption = "Main functions available in climatrends.") %>% 
  kable_styling(latex_options = "scale_down") %>% 
  column_spec(2, width = "1in") %>% 
  column_spec(4, width = "1.7in")
```

@@

>>>>

---
title: 'Climate variability indices for ecological and crop models in R: the `climatrends` package'
tags:
- climate data
- climatology
- earth science
- evapotranspiration
- precipitation data
- R
- reproducibility
- weather data
authors:
  - name: Kauê de Sousa
    orcid: 0000-0002-7571-7845
    affiliation: "1, 2"
  - name: Jacob van Etten
    orcid: 0000-0001-7554-2558
    affiliation: 2
  - name: Svein Ø. Solberg
    orcid: 0000-0002-4491-4483
    affiliation: 1
affiliations:
  - name: Department of Agricultural Sciences, Inland Norway University of Applied Sciences, Hamar, Norway
    index: 1
  - name: Bioversity International, Rome, Italy
    index: 2
citation_author: de Sousa et. al.
date: "`r format(Sys.time(), '%d %B %Y')`"
year: 2020
bibliography: paper.bib
csl: citationstyle.csl
output:
  pdf_document:
    keep_tex: false
header-includes: \usepackage{caption} \captionsetup[figure]{labelformat=empty}
---

```{r get data, include=FALSE}
library("knitr")
library("kableExtra")
library("tidyverse")
library("climatrends")
library("PlackettLuce")
load("cbean.rda")
```

@@

Abiotic factors play an important role in most ecological and crop systems that depend on certain levels of temperature, light and precipitation (and their interplay) to initiate important physiological events [@PlantEcology]. In the walk of climate change, understanding how these factors drive the physiological processes is a key approach to provide recommendations for adaptation and biodiversity conservation.

@@

`climatrends` aims to provide the R [@RCoreTeam] toolkit to compute extreme precipitation and temperature indices that serve as input for climate and crop models [@vanEtten2019; @Kehel2016], trends in climate change [@Aguilar2005; @deSousa2018] and applied ecology [@Prentice1992; @YLiu2018].

@@

Six main functions are provided, `crop_sensitive()`, `ETo()`, `GDD()`, `late_frost()`, `rainfall()` and `temperature()` with a default method for numeric 'vector' and additional methods implemented via the package `methods` [@RCoreTeam] for classes 'matrix' (or array), 'data.frame', and 'sf' (of geometry POINT or POLYGON) [@sf]. The last two methods are designed to fetch data from cloud sources.

@@

The indices computed by the functions `temperature()` and `rainfall()` are described in Table 2.

@@

where $T_{max}$ is the maximum temperature in the given day, $T_{min}$ is the minimum temperature in the given day and $T_{0}$ is the minimum temperature for growth (as per the physiology of the focal organism or ecosystem averages).

Additionally, the function `GDD()` offers three modified equations designed for cold environments and for tropical environments. For cold environments, where $T_{min}$ may be lower than $T_{0}$, there are two modified equations that adjust either $T_{mean}$ (variant a) or $T_{min}$ (variant b). The variant a changes $T_{mean}$ to $T_{0}$ if $T_{mean} < T_{0}$ and is expressed as follow.

@@

where $T$ may refer to $T_{min}$ and/or $T_{max}$ when the condition of being below $T_{0}$ applies.

@@

where $T_{0_{max}}$ is the maximum base temperature for growth, defined in `GDD()` using the argument `tbase_max`.

@@

where $GDD_{r}$ is the length of the vector with accumulated degree-days from day 1 to $n$.

@@

Late-spring frost is a freezing event occurring after a substantial accumulation of warmth. Frost damage is a known issue in temperate and boreal regions, it is associated with the formation of extracellular ice crystals that cause damage in the membranes [@Lambers2008]. Freezing occurring after an advanced phenological stage during spring may harm some plant species, resulting in lost of productivity in crop systems [@Trnka2014] and important ecological impacts [@Zohner2020].

The function `late_frost()` supports the computation of late-spring frost events. The function counts for the number of freezing days with minimum temperature below a certain threshold (argument `tfrost`). And returns the number of days spanned by frost events (temperature below `tfrost`), latency (event with no freezing temperature but also no accumulation of growing degree-days) and warming (when growing degree-days are accumulated enabling the development of the target organism). Additionally the function returns the first day of the events. The function calculates the growing degree-days applying the variant b (Eq. 3), which can be adjusted using the argument `equation` passed to `GDD()` as explained in the later section. The main inputs are a vector with maximum and minimum temperatures to compute the degree-days, a vector of dates (argument `date`), and, if needed, the `tbase` and `tfrost`, set by default to 4 and -2 $^\circ C$.

@@

Two functions in **climatrends** are mainly designed to capture the effects of climate on the development and stress of crop species, `crop_sensitive()` computes indices that aim to capture the changes in temperature extremes during key phenological stages (e.g. anthesis), and `ETo()` computes the reference evapotranspiration.

The crop sensitive indices available in **climatrends** are described in Table 3. These indices were previously used in crop models to project the impacts of climate change on crop yield [@Challinor2016; @Trnka2014]. Each index has a default temperature threshold(s) which can be adjusted by using the arguments `*.threshold`. Where the `*` means the index. For example, to change the defaults for hts_max (high temperature stress), a vector with the temperature thresholds is passed through the argument `hts_max.thresholds`.

The reference evapotranspiration measures the influence of the climate on a given plant's water need [@Brouwer1986]. The function `ETo()` applies the Blaney-Criddle method, a general theoretical method used when only air-temperature is available locally. It should be noted that this method is not very accurate and aims to provide the order of magnitude of evapotranspitation. The reference evapotranspiration is calculated using the following equation.

@@

Use the data from seed germination or crop growth to compute GDD. Maybe a GDD analysis in Scandinavia with data from the reindeer project.

## Common bean

During five growing seasons (from 2015 to 2017) in Nicaragua, van Etten et al. [-@vanEtten2019] conducted a crowdsourcing citizen-science experiment testing 11 common bean varieties (*Phaseolus vulgaris* L.) in 842 farmer-managed plots. Sets of three varieties were allocated randomly to farms as incomplete blocks. A Plackett–Luce model was used to analyse the data, this model estimates for each variety the probability that it wins, beating all other varieties in the set [@Turner2020]. An earlier version of `climatrends` was used in this research to capture the seasonal climate variation, here we reproduce part of this analysis regarding calculation and application of the climate indices. The approach here is slightly different because it considers the growing-degree days from planting date to maturity (the earlier study used planting date to the end of reproductive stage) and add new indices to illustrate the package implementation.

The data used here is available as supplementary material (**zenodo ref**) as `cbean`. This contains a data.frame with a Plackett-Luce grouped rankings, the geographical coordinates of each sampled plot and the planting dates from where each farmer decided to start the experiment. The planting dates differ from each other in the same season. The temperature data used was the land surface temperature MODIS (MYD11A2) [@Wan2015] and is storaged as an array with two layers (1st for the day and 2nd for the night temperatures). Each column corresponds to the dates (from 2015-09-10 to 2017-06-09) and the rows corresponds to the rows in the `cbean` data.frame.

@@

@@

```{r table1, echo=FALSE, results="asis"}
tabl <- as.data.frame(matrix(
  c("crop_sensitive()", "Compute crop sensitive indices",
    "ETo()", "Reference evapotranspiration using the Blaney-Criddle method",
    "GDD()", "Compute growing degree-days",
    "late_frost()","Compute the occurrence of late-spring frost",
    "rainfall()","Precipitation indices",
    "temperature()","Temperature indices"),
  nrow = 6, ncol = 2, byrow = TRUE
))

names(tabl) <- c("Function", "Definition")

kable(tabl, 
      align = "l", 
      caption = "Main functions available in climatrends.") %>% 
  kable_styling(latex_options = "hold_position")
```

```{r table2, echo=FALSE, results="asis"}
index <- as.data.frame(matrix(
  c("maxDT", "Maximun day temperature", "°C",
    "minDT", "Minimum day temperature", "°C",
    "maxNT", "Maximun night temperature", "°C",
    "minNT", "Minimum night temperature", "°C",
    "DTR", "Diurnal temperature range (mean difference between DT and NT)", "°C",
    "SU", "Summer days, number of days with maximum temperature > 30 °C", "days", 
    "TR", "Tropical nights, number of nights with maximum temperature > 25 °C", "days",
    "CFD", "Consecutive frosty days, number of days with temperature < 0 °C", "days",
    "WSDI","Maximum warm spell duration, consecutive days with temperature > 90th percentile", "days",
    "CSDI","Maximum cold spell duration, consecutive nights with temperature < 10th percentile","days",
    "T10p","The 10th percentile of night tempeture","°C",
    "T90p","The 90th percentile of day tempeture","°C",
    "MLDS", "Maximum length of consecutive dry day, rain < 1 mm", "days",
    "MLWS", "Maximum length of consecutive wet day, rain >= 1 mm", "days",
    "R10mm", "Heavy precipitation days 10 >= rain < 20 mm", "days",
    "R20mm", "Very heavy precipitation days rain >= 20", "days",
    "Rx1day", "Maximum 1-day precipitation", "mm",
    "Rx5day", "Maximum 5-day precipitation", "mm",
    "R95p", "Total precipitation when rain > 95th percentile", "mm",
    "R99p", "Total precipitation when rain > 99th percentile", "mm",
    "Rtotal", "Total precipitation in wet days, rain >= 1 mm", "mm",
    "SDII", "Simple daily intensity index, total precipitation divided by the number of wet days", "mm/days"),
  nrow = 22, ncol = 3, byrow = TRUE
))

names(index) <- c("Index", "Definition", "Unit")

kable(index,
      caption = "Temperature and precipitation indices available in climatrends.") %>% 
  kable_styling(latex_options = "hold_position") %>% 
  column_spec(2, width = "20em")
```

```{r cropsentbl, echo=FALSE, results="asis"}
index <- as.data.frame(matrix(
  c("hts_mean", "High temperature stress using daily mean temperature, and given as percentage number of days a certain threshold is exceeded", "32, 35, 38 °C",
    "hts_max", "High temperature stress using daily max temperature, and given as percentage number of days a certain threshold is exceeded", "36, 39, 42 °C",
    "hse", "Heat stress event, and given as percentage number of days a certain threshold is exceeded for at least two consecutive days", "31 °C", 
    "hse_ms", "Heat stress event, and given the maximum number of days a certain threshold is exceeded for at least two consecutive days", "31 °C",
    "cdi_mean", "Crop duration index using daily mean temperature, and given as max(Tmean - threshold, 0)", "22, 23, 24 °C",
    "cdi_max", "Crop duration index using daily max temperature, and given as max(Tmax - threshold, 0)", "27, 28, 29 °C",
    "lethal", "Lethal temperatures, defined as percentage of days during the timeseries where daily mean temperature exceeds a given threshold", "43, 46, 49 °C"),
  nrow = 7, ncol = 3, byrow = TRUE
))

names(index) <- c("Index", "Definition", "Default thresholds")

kable(index,
      caption = "Crop sensitive indices computed by climatrends.") %>% 
  kable_styling(latex_options = "hold_position") %>% 
  column_spec(2, width = "20em")

```
